---
title: 'P4-An Economic Entomology Example: The Spider Mite Data Set'
author: "David rt Fraser"
date: "January 21, 2016"
output: html_document
---
******
Steve Marchall of the Environmental Biology department informs me that the two-spotted spider mite is a major pest because it feeds on the leaves of strawberry plats, causing loss of chlorophyll, destruction of leaf tissue, and plant stunting.  The result is a loss of yield quality as well as quantity.  The subject of this study was to investigate the effect of a certain chemical agent designed to suppress the mite population.  Suppose the experiment was conducted as follows: each level of chlorinated hydrocarbon dosage was applied to 3 sets of 10 plants each which had been infested initially by mites.  The mites were brushed from the leaved onto glass plates and counted under magnification.  Below, Y represents the mean mite densities on a per-leaf basis and X represents the dosage level of the agent.  

******  

###**1.(a) What is the estimated regression line?**
```{r}
setwd("~/Documents/StudentUndergraduate/STAT 2050 -Design & Analysis of Experiments/1- Regression & Correlation Analysis/Assignment")
library(xlsx)
Spider<-read.xlsx(file="P4-Spider Mite Data.xlsx", sheetIndex = 1, header=TRUE)
Spider<-Spider[,-1]
names(Spider)<-c(10,20,30,40,50,75)
Spider
library(reshape2)
Spider<-melt(Spider)
Spider$variable<-as.integer(Spider$variable)*10
Spider$variable[16:18]<-75
model<-lm(Spider$value~Spider$variable)
model
```
The estimated regression line is $\hat{Y}$=`r round(model$coefficients[1],3)``r round(model$coefficients[2],3)`$\cdot{X}$.

###**(b) Complete the lack of fit ANOVA table.**
```{r}
anova(model)
lofmodel<-lm(Spider$value~as.factor(Spider$variable))
anova(model,lofmodel)
```
###**(c) Construct a scatter diagram and superimpose the estimated regression line.**
```{r}
plot(Spider$variable,Spider$value, main="Spider Mite Data Set", xlab="Dosage Level of Agent", ylab="Mean Mite Density/Leaf")
abline(model, col="blue")
```


