---
title: 'P7-An Aquatic Toxicology Example: The Brook Trout Data Set'
author: "David rt Fraser"
date: "January 22, 2016"
output: html_document
---
******
Bill Beamish of the University of Guelph Zoology department recently conducted an experiment to determine whether there was a significant relationship between the concentration of chromic oxide found in the fecal material of experimental brook trout, *Salvelinus fontinalis*, and the predetermined concentration levels of copper to which the fish were exposed. (If you wish to learn more about suck experiments, see Lett, P.F., Farmer, G.J. and Beamish, F.W.H. (1976). Effect of copper on some aspects of the bioenergetics of rainbow trout.  Fisheries Research Board of Canada, 33(6):1335-1242.)   

******
```{r}
setwd("~/Documents/StudentUndergraduate/STAT 2050 -Design & Analysis of Experiments/1- Regression & Correlation Analysis/Assignment")
library(xlsx)
Trout<-read.xlsx("P7-The Brook Trout Data Set.xlsx", sheetIndex = 1, header=TRUE)
Trout<-Trout[,-1]
names(Trout)<-c(0.000,0.075, 0.150, 0.225, 0.315)
suppressMessages(library(reshape2))
Trout<-melt(Trout)
Trout$variable<-as.numeric(Trout$variable)
Trout$variable <- gsub(1,0.000,Trout$variable)
Trout$variable <- gsub(2,0.075,Trout$variable)
Trout$variable <- gsub(3,0.150,Trout$variable)
Trout$variable <- gsub(4,0.225,Trout$variable)
Trout$variable <- gsub("^5$",0.315,Trout$variable)
Trout$variable <- as.numeric(Trout$variable)
Trout
model<-lm(Trout$value~Trout$variable)
model
```

###**(a) The estimated regression line is:**
$\hat{Y}$= `r round(model$coefficients[1],3)`+`r round(model$coefficients[2],3)`$\cdot{X}$  

###**(b) The correlation coefficient is:**
r = `r round(sqrt(summary(model)$"r.squared"),3)`

###**(c) Complete the ANOVA table.**
```{r}
anova(model)
lofmodel<-lm(Trout$value~as.factor(Trout$variable))
anova(model,lofmodel)
```

###**(d) Construct a scatter diagram and superimpose the estimated regression line.**
```{r}
plot(Trout$variable,Trout$value, main="Aquatic Toxicology: Brook Trout & Copper", xlab="Copper Conc. (mg/l)", ylab="Concentration of Chromic Oxide (%)")
abline(model, col="blue")
```  

###**(e) What are your recommendations to Professor Beamish about the relationship?**
Based on the results of the first ANOVA table F-value of 594.87, I would suggest that a significant correlation exists between the concentration of chromic oxide found in the fecal material of brook trout and the concentration level of copper to which those fish were exposed.  However, since the lack of fit ANOVA has a significant F-value (10.524), we would reject $H_O$, which implies that the straight line model is not a good fit.
